Make maps of loan words

### 2025-03-25 analysis/paper/paper.qmd
updates 24-3

# Apply the join function to each data frame in the list
data_list <- purrr::map(data_list, join_geometry)
```

The data may contain certain duplicate entries. These need to be removed to avoid overly full plots. This is done by running the code below.
```{r clean duplicates}
# Function to remove duplicates
remove_duplicates <- function(df_list) {
  df_list <- lapply(df_list, function(df) {
    df <- df |> 
      dplyr::distinct(glott_code, form, .keep_all = TRUE)
    return(df)
  })
  return(df_list)
}

@@

@@

Function to save a version of every map in the list of data
```{r save function}
# Function to create and save leaflet maps for all data frames in data_list
save_maps <- function(data_list, form_column = "form", background = "colour") {

@@

Save an html version of all objects
```{r}
#| label: save-maps
#| eval: false
#save_maps(data_list, background = "bw")

#create_leaflet_map(data_list[[2]])
```

>>>>

# Apply the join function to each data frame in the list
data_list <- purrr::map(data_list, join_geometry)

# Assign names
names(data_list) <- c("gold", "fonio", "millet", "salt", "iron")
```

The data may contain certain duplicate entries. These need to be removed to avoid overly full plots. This is done by running the code below.

```{r clean duplicates}
# Function to remove duplicates
remove_duplicates <- function(df_list) {
  df_list <- lapply(df_list, function(df) {
    df <- df |> 
      dplyr::distinct(glott_code, form, .keep_all = TRUE)
    return(df)
  })
  return(df_list)
}

@@

Classifications go into more detail than we need for plotting. We will only retain a rough classification here:

```{r}
data_list <- data_list |> 
  purrr::map(~ dplyr::mutate(.x, style = stringr::str_extract(style, "^[A-Za-z]")))
```

## Make interactive maps

@@

Function to save a version of every map in the list of data

```{r save function}
# Function to create and save leaflet maps for all data frames in data_list
save_maps <- function(data_list, form_column = "form", background = "colour") {

@@

```{r}
#| label: fonio
#| fig-cap: "fonio word forms"
 create_leaflet_map(data_list[[1]], background = "bw")
```

Save an html version of all objects

```{r}
#| label: save-maps
#| eval: false
save_maps(data_list)
```

## Make static maps (for printing etc)

Function to make static plots

```{r}
create_static_map <- function(data_sf, form_column = "form", style_column = "style", plot_title = NULL) {

  # Define the color palette for the "style" column, ensuring darker colors
  num_categories <- length(unique(data_sf[[style_column]]))
  dark_palette <- scales::brewer_pal(palette = "Dark2")(num_categories)

  # Create the base map with background
  base_map <- ggplot2::ggplot() +
    ggspatial::annotation_map_tile(tiles = "http://server.arcgisonline.com/ArcGIS/rest/services/Canvas/World_Light_Gray_Base/MapServer/tile/${z}/${y}/${x}") +
    ggplot2::geom_sf(data = data_sf, ggplot2::aes(color = !!rlang::sym(style_column)), show.legend = FALSE) +
    ggplot2::scale_color_manual(values = dark_palette) +
    ggplot2::theme_minimal() +
    ggplot2::theme(legend.position = "bottom") +
    ggplot2::labs(color = style_column)

  # Add the plot title if provided
  if (!is.null(plot_title)) {
    base_map <- base_map + ggplot2::ggtitle(plot_title)
  }

  # Crop the extent of the map to the data layers' extent
  base_map <- base_map + ggplot2::coord_sf(xlim = sf::st_bbox(data_sf)[c("xmin", "xmax")],
                                           ylim = sf::st_bbox(data_sf)[c("ymin", "ymax")])

  # Plot the data as strings contained in the form_column variable
  map_plot <- base_map +
    ggplot2::geom_text(data = data_sf, ggplot2::aes(x = sf::st_coordinates(data_sf)[,1],
                                                    y = sf::st_coordinates(data_sf)[,2],
                                                    label = !!rlang::sym(form_column),
                                                    color = !!rlang::sym(style_column)),
                       position = ggplot2::position_jitter(width = 0.1, height = 0.1), size = 3)

  # Return the ggplot object
  return(map_plot)
}
```

Make static plots

```{r}
create_static_map(data_list[[2]], plot_title = "fonio")
```

```{r}
create_static_map_tmap <- function(data_sf, form_column = "form", style_column = "style", plot_title = NULL) {

  # Define the color palette for the "style" column, ensuring darker colors
  num_categories <- length(unique(data_sf[[style_column]]))
  dark_palette <- scales::brewer_pal(palette = "Dark2")(num_categories)

  # Get the background map using maptiles package
  background_map <- maptiles::get_tiles(data_sf, provider = "Esri.WorldGrayCanvas", crop = TRUE)

  # Increase the boundaries of the plot
  bbox <- sf::st_bbox(data_sf)
  x_range <- bbox["xmax"] - bbox["xmin"]
  y_range <- bbox["ymax"] - bbox["ymin"]
  bbox["xmin"] <- bbox["xmin"] - 50 * x_range
  bbox["xmax"] <- bbox["xmax"] + 50 * x_range
  bbox["ymin"] <- bbox["ymin"] - 50 * y_range
  bbox["ymax"] <- bbox["ymax"] + 50 * y_range

  # Create the tmap object
  tmap::tmap_mode("plot")
  map_plot <- tmap::tm_shape(background_map) +
    tmap::tm_rgb() +
    tmap::tm_shape(data_sf) +
    tmap::tm_text(text = form_column, col = style_column, palette = dark_palette) +
    tmap::tm_layout(title = plot_title, legend.show = FALSE, bbox = bbox)

  # Return the tmap object
  return(map_plot)
}
```

```{r}
create_static_map_tmap(data_list[[1]])
```

## Make a network

Our network will be made using languages as nodes, which come with locations. We will then use the data in the list to create edges between the nodes.

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)

# Function to create edge list for a dataframe
create_edge_list <- function(df, df_name) {
  # Clean and transform the data
  df <- df |> 
    select(glott_code, style) |> 
    as_tibble()

  # Create edge list based on first letter of style
  edge_list_first_letter <- df |>
    mutate(first_letter = substr(style, 1, 1)) |>
    inner_join(., ., by = "first_letter") |>
    filter(glott_code.x != glott_code.y) |>
    select(glott_code.x, glott_code.y) |>
    distinct() |>
    mutate(type = df_name)

  # Create edge list based on first three characters of style
  edge_list_first_three <- df |>
    mutate(first_three = substr(style, 1, 3)) |>
    inner_join(., ., by = "first_three") |>
    filter(glott_code.x != glott_code.y) |>
    select(glott_code.x, glott_code.y) |>
    distinct() |>
    mutate(type = df_name)

  # Combine both edge lists
  combined_edge_list <- bind_rows(edge_list_first_letter, edge_list_first_three) |>
    distinct()

  return(combined_edge_list)
}

# Apply function to all dataframes in the list and combine results
all_edge_lists <- bind_rows(
  lapply(names(data_list), function(df_name) {
    create_edge_list(data_list[[df_name]], df_name)
  })
)

```

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)

# Subset the data
data_list_2 <- purrr::map(data_list, ~ {
  # Select the columns glott_code and style and convert to tibble
selected_df <- .x %>%
    as_tibble() %>%   # Convert to tibble
    select(glott_code, style) %>% # Select the relevant columns
    drop_na(style)

  return(selected_df)
})
```

```{r}

# Function to create edge list for a dataframe
create_edge_list <- function(df, df_name) {
  # Create edge list based on first letter of style
  df_first_letter <- df |>
    mutate(first_letter = substr(style, 1, 1))
  edge_list_first_letter <- df_first_letter |>
    inner_join(df_first_letter, by = "first_letter") |>
    filter(glott_code.x != glott_code.y) |>
    select(glott_code.x, glott_code.y) |>
    distinct() |>
    mutate(type = df_name, weight = 1)

  # Create edge list based on first three characters of style
  df_first_three <- df |>
    mutate(first_three = substr(style, 1, 3))
  edge_list_first_three <- df_first_three |>
    inner_join(df_first_three, by = "first_three") |>
    filter(glott_code.x != glott_code.y) |>
    select(glott_code.x, glott_code.y) |>
    distinct() |>
    mutate(type = df_name, weight = 1)

  # Combine both edge lists
  combined_edge_list <- bind_rows(edge_list_first_letter, edge_list_first_three) |>
    distinct()

  return(combined_edge_list)
}

# Apply function to all dataframes in the list and combine results
all_edge_lists <- bind_rows(
  lapply(names(data_list_2), function(df_name) {
    create_edge_list(data_list_2[[df_name]], df_name)
  })
)

# Summarize duplicates and calculate the weight
all_edge_lists_summarized <- all_edge_lists |>
  group_by(glott_code.x, glott_code.y, type) |>
  summarize(weight = sum(weight), .groups = 'drop')
```

Now, add the edge list and the node list to a tidygraph network

```{r}
library(tidygraph)

g <- tbl_graph(edges = all_edge_lists_summarized, directed = FALSE)

plot(g)
```
